---
title: "DATA 607 Project 1: Chess Tournament Analysis"
author: "Renee Watson"
date: "2026-09-16"
output: html_document
---

# Chess Tournament Analysis 

## Approach 

In this assignment, I will use a text file containing the results of a chess tournament with 64 players and seven rounds. The dataset includes each player's name, state, total points, pre-tournament rating, and results from each round. My plan is to load the text file into R and clean the data so that the information for each player can be organized into a structured dataset. 

One challenge I anticipate is that the original data is stored in a text format rather than a traditional table, with each player's information spread across multiple lines. I will need to separate the player information and extract the opponent numbers from the round results. I will then use the opponent numbers to match each opponent with their pre-tournament rating and calculate the average opponent pre-rating for each player. 

The final dataset will contain the player's name, state, total points, pre-tournament rating, and average pre-tournament rating of their opponents. I will export the cleaned dataset as a CSV file that can be imported into a SQL database. 

Data source: tournamentinfo.txt

## Codebase

### Load the Data 
```{r}
library(tidyverse)

chess <- readLines("tournamentinfo.txt")

head(chess, 10)
```

### Clean the Tournament Data 
```{r}
chess_data <- chess[-c(1:4)]

chess_data <- chess_data[!grepl("^---", chess_data)]

head(chess_data)
length(chess_data)
```

### Separate Player and Rating Information
```{r}
player_lines <- chess_data[seq(1, length(chess_data), by = 2)]

rating_lines <- chess_data[seq(2, length(chess_data), by = 2)]

head(player_lines)
head(rating_lines)
```


```{r}
player_split <- strsplit(player_lines, "\\|")

player_split[[1]]
```

### Extract Player Information
```{r}
player_number <- as.numeric(trimws(sapply(player_split, '[', 1)))
player_name <- trimws(sapply(player_split, '[', 2))
total_points <- as.numeric(trimws(sapply(player_split, '[', 3)))

head(player_number)
head(player_name)
head(total_points)
```


```{r}
rating_split <- strsplit(rating_lines, "\\|")
state <- trimws(sapply(rating_split, '[', 1))

head(state)
```

### Extract Pre-Tournament Ratings
```{r}
rating_info <- trimws(sapply(rating_split, '[', 2))
pre_rating <- as.numeric(
  sub(".*R:\\s*([0-9]+).*", "\\1", rating_info)
)

head(pre_rating)
```

### Create Player Dataset
```{r}
players <- data.frame(
  Player_Number = player_number,
  Player_Name = player_name,
  State = state,
  Total_Points = total_points,
  Pre_Rating = pre_rating
)

head(players)
```

### Extract Opponents
```{r}
rounds <- sapply(player_split, function(x) x[4:10])

rounds <- t(rounds)

head(rounds)
```


```{r}
opponents <- apply(rounds, c(1,2), function(x) {
  number <- gsub("[^0-9]", "", x)
  ifelse(number == "", NA, as.numeric(number))
}
  )

head(opponents)
```

### Calculate Average Opponent Ratings
```{r}
avg_opponent_rating <- apply(opponents, 1, function(x) {
  x <- x[!is.na(x)]
  opponent_ratings <- players$Pre_Rating[
    match(x, players$Player_Number)
  ]
  mean(opponent_ratings)
}
  )

head(avg_opponent_rating)
```


```{r}
players$Avg_Opponent_Pre_Rating <- round(avg_opponent_rating)
head(players)
```

### Create Final Dataset
```{r}
final_data <- players %>%
  select(
    Player_Name,
    State,
    Total_Points,
    Pre_Rating,
    Avg_Opponent_Pre_Rating
  )

head(final_data)
nrow(final_data)
```


```{r}
write.csv(
  final_data,
  "chess_tournament_results.csv",
  row.names = FALSE
)

file.exists("chess_tournament_results.csv")
```


## Conclusion

In this project, I transformed the chess tournament text file into a structured dataset containing information for all 64 players. I extracted each player's name, state, total points, and pre-tournament rating. I also used the results from each round to identify each player's opponents and calculate the average pre-tournament rating of those opponents. The main challenge was working with the semi-structured text file because the player information was spread across multiple lines and the round results contained both letters and opponent numbers. After cleaning and organizing the data, I was able to create the five required variables and export the final dataset as a CSV file that can be imported into a SQL database. 














